---
title: "morphStuart"
author: "Aaron Myrold"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
#==============================================================================#
# PROJECT INTRODUCTION

#==============================================================================#
# SETUP
## Loading Libraries
```{r Libraries, include=FALSE}
# load libraries needed for our analyses
# 1 ggplot2
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

# 2 dplyr
if (!require("dplyr")) install.packages("dplyr")
library(dplyr)

# 3 tidyr
if (!require("tidyr")) install.packages("tidyr")
library(tidyr)

# 4 stringr
if (!require("stringr")) install.packages("stringr")
library(stringr)

```

## Global Variables
```{r Global Variables, include=FALSE}
# store current user's working directory
wk.dir <- getwd()
# Set seed for reproducibility
set.seed(333)

```

## Folder Management
```{r Folder Management, include=FALSE}

# This is a function to initialize a new project
  # It creates required subfolders (see defaults) 
  # It outputs an object that stores the relative paths to these sub-folders

# If calling the function within the working directory of the project, project_number = 0
  # this will create necessary sub-folders in current working directory

# If creating a new root folder for the project, provide project_number >=1 and a project_name
  # this will create a folder "pX.project_name/" to store all sub-folders
new_project <- function(project_number, project_name = NULL, subfolders = NULL, wk.dir = getwd()) {
  # Set default subfolders
  default_subfolders <- c("data", "scripts", "output", "docs")
  
  # Use provided sub-folders or defaults if none are provided
  if (is.null(subfolders)) {
    subfolders <- default_subfolders
  }
  
  # Initialize project paths list
  project_paths <- list()
  
  # Check if this is a "root directory" project (project_number = 0)
  is_root_project <- project_number == 0
  
  if (is_root_project) {
    # Use working directory as the base path for root projects
    base_path <- wk.dir
    project_paths$main <- wk.dir
    message("Creating folders directly in working directory: ", wk.dir)
  } else {
    # For normal projects, validate project name is provided
    if (is.null(project_name)) {
      stop("Project name is required when project_number is not 0")
    }
    
    # Create project folder name
    project_folder <- paste0("p", project_number, ".", project_name)
    
    # Full path for the project folder
    project_path <- file.path(wk.dir, project_folder)
    
    # Create the project folder if it doesn't exist
    if (!file.exists(project_path)) {
      dir.create(project_path)
      message("Created project folder: ", project_path)
    } else {
      message("Project folder already exists: ", project_path)
    }
    
    # Set base path for subfolders
    base_path <- project_path
    project_paths$main <- project_path
  }
  
  # Create sub-folders
  for (sub in subfolders) {
    # Determine subfolder name based on project type
    if (is_root_project) {
      subfolder_name <- sub
    } else {
      subfolder_name <- paste0("p", project_number, ".", sub)
    }
    
    # Full path for the subfolder
    subfolder_path <- file.path(base_path, subfolder_name)
    
    # Create sub-folder if it doesn't exist
    if (!file.exists(subfolder_path)) {
      dir.create(subfolder_path, recursive = TRUE)  # Added recursive=TRUE for nested paths
      message("Created subfolder: ", subfolder_path)
    } else {
      message("Subfolder already exists: ", subfolder_path)
    }
    
    # Add the sub-folder path to the list
    project_paths[[sub]] <- paste0(subfolder_path, "/")
  }
  
  # Return the list of paths as an object
  return(project_paths)
}

```

## Initialize Directories
```{r}
# Initialize new project directories
p <- new_project(0, subfolders = c("1.data", "1.data/a.raw", "1.data/b.merged",
                                   "1.data/c.flagged","1.data/d.processed", 
                                   "2.scripts", "3.outputs", "4.docs"))
```

#==============================================================================#
# FUNCTIONS
## variable_mapping
```{r}
variable_mapping <- function() {
  # Define measurement types
  continuous_vars <- c("SL", "CAV", "DS1", "DS2", "DS3", "LPT", 
                      "PSP.L", "PSP.R", "TPG", "ECT", "CLE", "PMX")
  
  count_vars <- c("MDF", "MAF", "MCV", "MAV", "MPT", "MPSP")
  
  binary_vars <- c("MDS1", "MDS2", "MDS3", "MDS1NA", 
                   "MDS2NA", "MDS3NA", "MPSPNA")
  
  # Return as a list
  return(list(
    continuous = continuous_vars,
    count = count_vars,
    binary = binary_vars,
    all = c(continuous_vars, count_vars, binary_vars)
  ))
}
```

## handle_special_cases
```{r}
handle_special_cases <- function(data) {
  # requires dplyr
  
  # Use dplyr to handle all the special cases with cleaner syntax
  cleaned <- data %>%
    as_tibble() %>%
    # Handle dorsal spine columns - use case_when for readable conditional logic
    mutate(
      # Dorsal spine 1
      DS1 = case_when(
        MDS1 == 0 & MDS1NA == 1 ~ NA_real_,
        MDS1 == 0 & MDS1NA == 0 & is.na(DS1) ~ 0,
        TRUE ~ DS1
      ),
      # Dorsal spine 2
      DS2 = case_when(
        MDS2 == 0 & MDS2NA == 1 ~ NA_real_,
        MDS2 == 0 & MDS2NA == 0 & is.na(DS2) ~ 0,
        TRUE ~ DS2
      ),
      # Dorsal spine 3
      DS3 = case_when(
        MDS3 == 0 & MDS3NA == 1 ~ NA_real_,
        MDS3 == 0 & MDS3NA == 0 & is.na(DS3) ~ 0,
        TRUE ~ DS3
      ),
      
      # Handle MPT (pre-dorsal pterygiophores)
      MPT = case_when(
        MPT == 0 & (MDS1 == 1 | MDS2 == 1 | MDS3 == 1) ~ NA_real_,
        TRUE ~ MPT
      ),
      
      # Handle pelvic spine columns
      PSP.L = case_when(
        MPSP == 0 & MPSPNA == 1 ~ NA_real_,
        TRUE ~ PSP.L
      ),
      PSP.R = case_when(
        MPSP == 0 & MPSPNA == 1 ~ NA_real_,
        TRUE ~ PSP.R
      )
    )
  
  # When TPGNA column is available, apply similar logic to TPG
  if ("TPGNA" %in% names(cleaned)) {
    cleaned <- cleaned %>%
      mutate(
        TPG = case_when(
          TPGNA == 1 ~ NA_real_,
          TRUE ~ TPG
        )
      )
  }
  
  return(as.data.frame(cleaned))
}

```

## identify_overlaps
```{r}
identify_overlaps <- function(data, threshold = 0.05) {
  # Requires tidyr and dplyr
  
  # Get variable mappings
  var_map <- variable_mapping()
  
  # Step 1: Create separate dataframes for part and counterpart data
  part_data <- data %>% 
    filter(part_type == "P") %>%
    select(fish_id, part_type, row_id = n, Scale_10mm, all_of(var_map$all))
  
  cpart_data <- data %>% 
    filter(part_type == "C") %>%
    select(fish_id, part_type, row_id = n, Scale_10mm, all_of(var_map$all))
  
  # Step 2: Gather measurements into long format for easier comparison
  part_long <- part_data %>%
    pivot_longer(
      cols = all_of(var_map$all),
      names_to = "measure",
      values_to = "part_value"
    ) %>%
    filter(!is.na(part_value)) %>%
    select(fish_id, part_row_id = row_id, measure, part_value, part_scale = Scale_10mm)
  
  cpart_long <- cpart_data %>%
    pivot_longer(
      cols = all_of(var_map$all),
      names_to = "measure",
      values_to = "cpart_value"
    ) %>%
    filter(!is.na(cpart_value)) %>%
    select(fish_id, cpart_row_id = row_id, measure, cpart_value, cpart_scale = Scale_10mm)
  
  # Step 3: Join part and counterpart data to find overlaps
  overlaps <- part_long %>%
    inner_join(cpart_long, by = c("fish_id", "measure"))
  
  # Step 4: Calculate differences based on variable type
  overlaps <- overlaps %>%
    mutate(
      var_type = case_when(
        measure %in% var_map$continuous ~ "continuous",
        measure %in% var_map$count ~ "count",
        measure %in% var_map$binary ~ "binary",
        TRUE ~ "other"
      ),
      scale = coalesce((part_scale + cpart_scale) / 2, part_scale, cpart_scale),
      # For continuous vars: use relative difference
      absolute_diff = abs(part_value - cpart_value),
      relative_diff = case_when(
        var_type == "continuous" ~ absolute_diff / scale,
        TRUE ~ as.numeric(absolute_diff > 0) # Binary indicator for non-continuous
      ),
      # Different flagging criteria based on variable type
      exceeds_threshold = case_when(
        var_type == "continuous" ~ relative_diff > threshold,
        var_type == "count" ~ absolute_diff > 0, # Any difference for counts
        var_type == "binary" ~ absolute_diff > 0, # Any difference for binary
        TRUE ~ FALSE
      )
    )
  
  # Step 5: Get list of fish with overlaps that meet criteria
  fish_with_significant_overlaps <- overlaps %>%
    filter(exceeds_threshold) %>%
    select(fish_id) %>%
    distinct() %>%
    pull(fish_id)
  
  # Step 6: Separate fish with and without significant overlaps
  overlap_fish <- data %>%
    filter(fish_id %in% fish_with_significant_overlaps)
  
  non_overlap_fish <- data %>%
    filter(!fish_id %in% fish_with_significant_overlaps)
  
  # Convert results back to standard data.frames for consistency
  return(list(
    overlap_fish = as.data.frame(overlap_fish),
    non_overlap_fish = as.data.frame(non_overlap_fish),
    overlap_metrics = as.data.frame(overlaps)
  ))
}

```
## merge_non_overlap
```{r}
merge_non_overlap <- function(data) {
  # requires dplyr and tidyr
  
  # Get unique fish IDs
  fish_ids <- unique(data$fish_id)
  
  # Define measurement columns
  id_cols <- c("n", "ID", "fish_id", "part_type")
  measure_cols <- setdiff(names(data), id_cols)
  
  # Initialize empty dataframe for results
  result_list <- list()
  
  # Process each fish - some looping is still needed for the logical merging
  for (i in seq_along(fish_ids)) {
    id <- fish_ids[i]
    
    # Get all data for this fish
    fish_data <- data %>% 
      filter(fish_id == id)
    
    # If only one row, just use it directly
    if (nrow(fish_data) == 1) {
      result_list[[i]] <- fish_data
      next
    }
    
    # Step 1: Convert to long format (each measurement becomes a row)
    fish_long <- fish_data %>%
      pivot_longer(
        cols = all_of(measure_cols),
        names_to = "variable",
        values_to = "value"
      ) %>%
      filter(!is.na(value))  # Remove NA values
    
    # Step 2: Prioritize part rows (P) over counterpart rows (C)
    merged_values <- fish_long %>%
      group_by(variable) %>%
      arrange(variable, ifelse(part_type == "P", 1, 2)) %>%  # Sort P before C
      slice(1) %>%  # Take first row in each group (prioritizes P)
      ungroup()
    
    # Step 3: Convert back to wide format
    merged_row <- merged_values %>%
      select(variable, value) %>%
      pivot_wider(
        names_from = "variable",
        values_from = "value"
      ) 
    
    # Step 4: Add back fish_id and set part_type to "merged"
    merged_row <- merged_row %>%
      mutate(
        fish_id = id,
        part_type = "merged",
        merged = TRUE
      )
    
    # Add to result list
    result_list[[i]] <- merged_row
  }
  
  # Combine all results
  result <- bind_rows(result_list)
  
  return(as.data.frame(result))
}

```

## create_review_list
```{r}
create_review_list <- function(overlap_results, threshold = 0.05) {
  # requires dplyr
  metrics <- overlap_results$overlap_metrics
  
  # Get variable mappings
  var_map <- variable_mapping()
  
  # Add more meaningful grouping and sorting
  summary <- metrics %>%
    # First summarize by fish and variable type
    group_by(fish_id, var_type) %>%
    summarize(
      has_diffs = any(exceeds_threshold),
      num_overlapping = n(),
      num_with_diffs = sum(exceeds_threshold),
      # For continuous vars, calculate max difference
      max_relative_diff = if_else(
        first(var_type) == "continuous", 
        max(relative_diff, na.rm = TRUE), 
        NA_real_
      ),
      .groups = "drop"
    ) %>%
    # Then create a summary by fish
    group_by(fish_id) %>%
    summarize(
      continuous_diffs = sum(var_type == "continuous" & has_diffs),
      count_diffs = sum(var_type == "count" & has_diffs),
      binary_diffs = sum(var_type == "binary" & has_diffs),
      total_diffs = sum(has_diffs),
      max_relative_diff = max(max_relative_diff, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    # Sort by total differences
    arrange(desc(total_diffs), desc(continuous_diffs))
  
  # Create a more detailed breakdown by measurement type
  details <- metrics %>%
    filter(exceeds_threshold) %>%
    arrange(fish_id, var_type, desc(relative_diff))
  
  return(list(
    summary = as.data.frame(summary),
    details = as.data.frame(details)
  ))
}

```

## export_review_list
```{r}
export_review_list <- function(review_results, file_prefix = "fish_review", 
                               output_dir = NULL) {
  # Create output directory if specified and it doesn't exist
  if (!is.null(output_dir)) {
    if (!dir.exists(output_dir)) {
      dir.create(output_dir, recursive = TRUE)
    }
    summary_path <- file.path(output_dir, paste0(file_prefix, "_summary.csv"))
    details_path <- file.path(output_dir, paste0(file_prefix, "_details.csv"))
  } else {
    summary_path <- paste0(file_prefix, "_summary.csv")
    details_path <- paste0(file_prefix, "_details.csv")
  }
  
  # Export files
  readr::write_csv(review_results$summary, summary_path)
  readr::write_csv(review_results$details, details_path)
  
  # Return file paths
  return(list(
    summary = summary_path,
    details = details_path
  ))
}

```

## validate_data
```{r}
# Add a validation function to ensure data quality
validate_data <- function(original, corrected, merged, overlaps) {
  # requires dplyr
  
  # Initialize results
  validation <- list()
  
  # Check if any rows were lost
  total_specimens_original <- length(unique(original$fish_id))
  total_specimens_after <- length(unique(c(
    merged$fish_id, 
    unique(overlaps$fish_id)
  )))
  
  validation$all_specimens_accounted_for <- 
    total_specimens_original == total_specimens_after
  
  validation$specimen_counts <- tibble(
    stage = c("Original", "Merged", "Flagged for review", "Total after processing"),
    count = c(
      total_specimens_original,
      length(unique(merged$fish_id)),
      length(unique(overlaps$fish_id)),
      total_specimens_after
    )
  )
  
  # Check special case handling
  special_case_check <- corrected %>%
    filter(
      (MDS1 == 0 & MDS1NA == 1 & !is.na(DS1)) |
      (MDS2 == 0 & MDS2NA == 1 & !is.na(DS2)) |
      (MDS3 == 0 & MDS3NA == 1 & !is.na(DS3)) |
      (MPT == 0 & (MDS1 == 1 | MDS2 == 1 | MDS3 == 1)) |
      (MPSP == 0 & MPSPNA == 1 & (!is.na(PSP.L) | !is.na(PSP.R)))
    )
  
  validation$special_cases_properly_handled <- nrow(special_case_check) == 0
  validation$problematic_rows <- special_case_check
  
  return(validation)
}

```

```{r}


```

```{r}


```

#==============================================================================#
# DATA IMPORT
```{r}
# Read in data and check initial structure
morph <- read.csv(paste0(p$`1.data/a.raw`, "241015_PitLMorph.csv"))
# str(morph)
```

## ID Structure Analysis
```{r}
# Extract fish ID and part type
morph <- morph %>%
  mutate(
    fish_id = str_remove(ID, "_[PC]\\.jpg$"),
    part_type = str_extract(ID, "[PC](?=\\.jpg$)")
  )

# Count rows per fish and part type
fish_part_counts <- morph %>%
  group_by(fish_id, part_type) %>%
  summarise(row_count = n(), .groups = "drop") %>%
  pivot_wider(names_from = part_type, values_from = row_count, values_fill = 0)

# Check fish without both part types
missing_parts <- fish_part_counts %>%
  filter(P == 0 | C == 0)
sum(missing_parts$`NA`) # 14 fish without both "part" & "counter part"

# Fish with excessive rows
multi_row_fish <- fish_part_counts %>%
  filter(P > 1 | C > 1)
length(multi_row_fish$fish_id) # 361 fish with duplicate "parts" or "counter parts"

```

## Basic Quality Check
```{r}
# Check for implausible values in measurement columns
# Example for length measurements (adjust for your specific columns)
length_cols <- c("SL", "CAV", "DS1", "DS2", "DS3", "LPT", "PSP.L", "PSP.R", "TPG", "ECT", "CLE", "PMX")

# Check for negative values in length columns
negative_lengths <- morph %>%
  select(ID, all_of(length_cols)) %>%
  pivot_longer(cols = -ID, names_to = "measure", values_to = "value") %>%
  filter(value < 0)
length(negative_lengths$ID) # No negative measurements
```

#==============================================================================#
# DATA PREPROCESSING
## Handle Special Cases
```{r Spine and Girdle Logic}
# Apply logic for dorsal spines and pelvic girdle:
# - If *NA = 1, corresponding measurements should be NA (poor preservation)
# - If *NA = 0 and respective count = 0, length measurements should be 0 (biologically absent)

morph_corrected <- handle_special_cases(morph)

```

## Overlap Assesment
```{r}
overlap_results <- identify_overlaps(morph_corrected, threshold = 0.05)

# Summarize overlap findings
cat("Number of fish with overlapping measurements:", 
    length(unique(overlap_results$overlap_metrics$fish_id)), "\n")
cat("Number of fish without overlaps:", 
    length(unique(overlap_results$non_overlap_fish$fish_id)), "\n")

```

## Check if Missing Scale
```{r}
# Found ~43 records with missing scale information.
# Below is code checking the source of these errors

# Check how many fish have missing scale information
scale_summary <- morph_corrected %>%
  group_by(fish_id) %>%
  summarize(
    has_part_scale = any(!is.na(Scale_10mm[part_type == "P"])),
    has_cpart_scale = any(!is.na(Scale_10mm[part_type == "C"])),
    has_any_scale = has_part_scale | has_cpart_scale,
    .groups = "drop"
  )

print(paste("Fish without any scale information:", 
            sum(!scale_summary$has_any_scale), 
            "out of", nrow(scale_summary)))

# Check specifically which problems are causing the non-finite relative_diff values
problem_rows <- overlap_results$overlap_metrics %>% 
  filter(!is.finite(relative_diff)) %>%
  mutate(
    problem_type = case_when(
      is.na(part_scale) & is.na(cpart_scale) ~ "both_scales_missing",
      is.na(part_scale) ~ "part_scale_missing",
      is.na(cpart_scale) ~ "cpart_scale_missing",
      scale == 0 ~ "zero_scale",
      is.na(part_value) | is.na(cpart_value) ~ "measurement_missing",
      TRUE ~ "other_issue"
    )
  )

# Summarize the types of problems
problem_summary <- problem_rows %>%
  group_by(problem_type) %>%
  summarize(count = n(), .groups = "drop")

print(problem_summary)

# Get list of fish IDs missing scales
fish_missing_scales <- scale_summary %>%
  filter(!has_any_scale) %>%
  pull(fish_id)

# Create report of fish missing scale information
missing_scales_detail <- morph %>%
  filter(fish_id %in% fish_missing_scales) %>%
  select(fish_id, ID, part_type, Scale_10mm) %>%
  arrange(fish_id, part_type)

# Write the detailed report
write.csv(missing_scales_detail, 
          file = paste0(p$`1.data/c.flagged`, "/fish_missing_scales_detailed.csv"), 
          row.names = FALSE)

```

#==============================================================================#
# DATA CLEANING
## Merge Non-Overlapping Fish
```{r}
merged_non_overlap <- merge_non_overlap(overlap_results$non_overlap_fish)
```

## Flag Conflicting Measurements
```{r Flag Duplicates}
review_results <- create_review_list(overlap_results)

export_review_list(review_results, file_prefix = "fish_fossil_review",
                           output_dir = p$`1.data/c.flagged`)
```

## Visualize Results
```{r}
# Visualization 1: Distribution of relative differences
ggplot(overlap_results$overlap_metrics, aes(x = relative_diff)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  geom_vline(xintercept = 0.05, color = "red", linetype = "dashed") +
  labs(title = "Distribution of Relative Differences",
       x = "Relative Difference (proportion of scale)",
       y = "Count") +
  theme_minimal()

# Visualization 2: Count of overlaps by measurement type
overlap_by_measure <- overlap_results$overlap_metrics %>%
  group_by(measure) %>%
  summarise(
    count = n(),
    avg_diff = mean(relative_diff, na.rm = TRUE),
    exceed_threshold = sum(exceeds_threshold),
    .groups = "drop"
  )

ggplot(overlap_by_measure, aes(x = reorder(measure, count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Number of fish with duplicate measurements by type",
       x = "Measurement",
       y = "Count of fish with duplicates")

# Visualization 3: Distribution of flags per fish
ggplot(review_results$summary, aes(x = total_diffs)) +
  geom_histogram(binwidth = 1, fill = "coral") +
  theme_minimal() +
  labs(title = "Distribution of duplicate measurements per fish",
       x = "Number of duplicate measurements",
       y = "Count of fish")

# Visualization 4: Distribution of differences by variable type
ggplot(overlap_results$overlap_metrics, 
       aes(x = var_type, y = absolute_diff, fill = var_type)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Distribution of Differences by Variable Type",
       x = "Variable Type", 
       y = "Absolute Difference",
       fill = "Variable Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

