---
title: "Stickleback Morphology & Paleo-Ecology Data Processing Pipeline"
author: "Analysis Report"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(targets)
library(dplyr)
library(knitr)
library(visNetwork)
library(rioja)

# Load key targets that affect the report (for dependency tracking)
tar_load(c(
  # Raw data
  "morph_raw", "paleo_raw", "order_raw",
  
  # Morphology pipeline
  "morph_with_ids", "morph_with_scales", "morph_corrected", 
  "morph_non_overlap", "morph_final", "flagged_fish_missing_scales",
  
  # Paleo pipeline  
  "paleo_with_ids", "paleo_merged_counts",
  
  # Field order processing
  "order_with_ids", "fieldorder_formatted", "fieldorder_duplicates_clean", 
  "fieldorder_updated", "fieldorder_final", "fieldorder_processing_summary",
  
  # Integration and stratigraphy
  "lspec_integration_check", "paleo_summary_table", "rioja_strat_plots",
  
  # Community ecology
  "community_metrics", "turnover_matrix",
  
  # Report plots
  "plot_completeness", "plot_microfossil_types", "trends_plot", "turnover_heatmap"
))

```

<div style="background-color: #f8f9fa; border: 2px solid #dee2e6; border-radius: 8px; padding: 20px; margin: 15px 0; max-width: 100%; overflow: hidden;">

# Pipeline Overview

This report documents the data processing pipeline for stickleback morphological measurements and paleo-ecological microfossil data. The pipeline transforms raw measurements into clean, analysis-ready datasets through systematic quality control and standardization steps.

## Full Pipeline Architecture

```{r pipeline-viz, fig.height=6, out.width="85%", fig.align="center"}
tryCatch({
  core_targets <- c(
    "morph_raw", "paleo_raw", "order_raw",
    "morph_with_ids", "paleo_with_ids", "order_with_ids",
    "morph_with_scales", "morph_corrected", "morph_non_overlap", "morph_final",
    "paleo_merged_counts", "fieldorder_formatted", "fieldorder_duplicates_clean", 
    "fieldorder_final", "paleo_stickleback_only", "lspec_integration_check", 
    "paleo_with_depths", "paleo_summary_table", "rioja_species_matrix", 
    "rioja_depth_data", "rioja_strat_plots"
  )
  
  tar_visnetwork(
    targets_only = TRUE, 
    names = core_targets,
    label = c("time", "size")
  ) %>%
    visLegend(zoom = FALSE)
}, error = function(e) {
  cat("Pipeline visualization not available.\n")
})
```

---

# Morphology Processing Pipeline

The morphology pipeline processes individual fish measurements from fossilized specimens, handling both "part" (P) and "counterpart" (C) sides of the same fossil. The goal is to create a single, high-quality measurement record for each fish.

## Morphology Pipeline Flow

```{r morph-pipeline-viz, fig.height=6}
tryCatch({
  morph_targets <- c(
    "morph_raw", "morph_with_ids", "flagged_fish_missing_scales", 
    "morph_with_scales", "morph_corrected", "morph_non_overlap", "morph_final"
  )
  
  tar_visnetwork(
    targets_only = TRUE, 
    names = morph_targets,
    label = c("time", "size")
  ) %>%
    visLegend(zoom = FALSE)
}, error = function(e) {
  cat("Morphology pipeline visualization not available.\n")
})
```

## Step 1: ID Extraction

The `extract_ids()` function parses complex specimen identifiers to extract key information for linking records across datasets.

**Input ID Format**: `V330363_L2_1_C.jpg`  
**Extracted Components**:
- `fish_id`: V330363 (individual fish specimen)
- `LSPEC`: L0002 (standardized geological specimen number)
- `part_type`: C (counterpart vs P for part)

```{r id-extraction}
tryCatch({
  cat("Raw data structure:\n")
  cat("Records:", nrow(morph_raw), "\n")
  cat("Unique IDs:", length(unique(morph_raw$ID)), "\n\n")
  
  cat("After ID extraction:\n")
  cat("Unique fish_id values:", length(unique(morph_with_ids$fish_id[!is.na(morph_with_ids$fish_id)])), "\n")
  cat("Unique LSPEC values:", length(unique(morph_with_ids$LSPEC[!is.na(morph_with_ids$LSPEC)])), "\n")
  
  # Show example of ID parsing
  sample_ids <- head(morph_with_ids[!is.na(morph_with_ids$fish_id), c("ID", "fish_id", "LSPEC", "part_type")], 5)
  kable(sample_ids, caption = "Example ID Parsing Results")
  
  # Part/Counterpart distribution
  part_dist <- table(morph_with_ids$part_type, useNA = "ifany")
  cat("\nPart/Counterpart Distribution:\n")
  print(part_dist)
}, error = function(e) {
  cat("ID extraction data not available.\n")
})
```

## Step 2: Scale Bar Quality Control

Scale bars (10mm reference) are essential for measurement standardization. The pipeline either incorporates updated scale data or removes fish lacking scale information.

```{r scale-handling}
tryCatch({
  cat("Scale Bar Quality Control:\n")
  cat("Fish missing scales:", length(flagged_fish_missing_scales), "\n")
  cat("Records before scale handling:", nrow(morph_with_ids), "\n")
  cat("Records after scale handling:", nrow(morph_with_scales), "\n")
  cat("Fish retained with scales:", length(unique(morph_with_scales$fish_id)), "\n\n")
  
  # Check if updated scales were incorporated
  if (file.exists("data/raw/160625_missingscales.txt")) {
    cat("Updated scale data was incorporated from colleague's measurements.\n")
  } else {
    cat("Fish without scale data were removed from analysis.\n")
  }
}, error = function(e) {
  cat("Scale handling data not available.\n")
})
```

## Step 3: Evolutionary Loss Logic

The `evolved_loss()` function applies biological logic to distinguish true evolutionary loss (coded as 0) from missing data due to poor preservation (coded as NA). This is critical for dorsal spine and pelvic girdle measurements.

**Biological Logic Applied**:
- If `MDS1 = 0` & `MDS1NA = 1`: Set `DS1` to `NA` (lost to preservation)
- If `MDS1 = 0` & `MDS1NA = 0`: Set `DS1` to `0` (true evolutionary loss)

```{r evolutionary-loss}
tryCatch({
  affected_cols <- c("DS1", "DS2", "DS3", "PSP.L", "PSP.R", "TPG")
  available_cols <- intersect(affected_cols, names(morph_corrected))
  
  if (length(available_cols) > 0) {
    cat("Evolutionary Loss Corrections Applied:\n\n")
    
    for (col in available_cols) {
      if (col %in% names(morph_with_scales)) {
        na_before <- sum(is.na(morph_with_scales[[col]]))
        na_after <- sum(is.na(morph_corrected[[col]]))
        zero_before <- sum(morph_with_scales[[col]] == 0, na.rm = TRUE)
        zero_after <- sum(morph_corrected[[col]] == 0, na.rm = TRUE)
        
        cat(sprintf("%s: %d → %d NAs (%+d), %d → %d zeros (%+d)\n", 
                   col, na_before, na_after, na_after - na_before,
                   zero_before, zero_after, zero_after - zero_before))
      }
    }
    cat("\nThis transformation preserves true evolutionary loss while properly coding preservation artifacts.\n")
  } else {
    cat("Evolutionary loss columns not found in dataset.\n")
  }
}, error = function(e) {
  cat("Evolutionary loss data not available.\n")
})
```

## Step 4: Conflict Detection and Removal

Fish with conflicting measurements between part (P) and counterpart (C) sides are flagged and removed. The threshold for conflicts is set at 5% relative difference for continuous measurements.

```{r conflict-removal}
tryCatch({
  cat("Conflict Detection and Removal:\n")
  cat("Records before conflict check:", nrow(morph_corrected), "\n")
  cat("Fish before conflict check:", length(unique(morph_corrected$fish_id)), "\n")
  cat("Records after conflict removal:", nrow(morph_non_overlap), "\n")
  cat("Fish after conflict removal:", length(unique(morph_non_overlap$fish_id)), "\n\n")
  
  # Calculate fish removed
  fish_before <- unique(morph_corrected$fish_id)
  fish_after <- unique(morph_non_overlap$fish_id)
  fish_removed <- setdiff(fish_before, fish_after)
  
  cat("Fish flagged for conflicts:", length(fish_removed), "\n")
  cat("Conflict rate:", round(100 * length(fish_removed) / length(fish_before), 1), "%\n\n")
  
  cat("Conflicting fish are saved to 'data/flagged/counterpart_conflicts.csv' for manual review.\n")
}, error = function(e) {
  cat("Conflict removal data not available.\n")
})
```

## Step 5: Part/Counterpart Merging

For fish without conflicts, part and counterpart measurements are intelligently merged, prioritizing part (P) measurements over counterpart (C) when both are available.

```{r merging-summary}
tryCatch({
  cat("Part/Counterpart Merging Results:\n")
  cat("Input records (P + C):", nrow(morph_non_overlap), "\n")
  cat("Output records (merged):", nrow(morph_final), "\n")
  cat("Final fish count:", length(unique(morph_final$fish_id)), "\n\n")
  
  if ("merged" %in% names(morph_final)) {
    merge_success <- sum(morph_final$merged, na.rm = TRUE)
    cat("Successfully merged records:", merge_success, "\n")
  }
}, error = function(e) {
  cat("Merging summary data not available.\n")
})
```

## Data Completeness

```{r completeness-plot, fig.height=6}
tryCatch({
  print(plot_completeness)
}, error = function(e) {
  cat("Completeness plot not available.\n")
})
```

---

# Paleo-Ecology Processing Pipeline

The paleo-ecology pipeline processes microscopy count data, where each row represents individual microscopy line counts that must be aggregated into geological sample totals.

## Paleo Pipeline Flow

```{r paleo-pipeline-viz, fig.height=4}
tryCatch({
  paleo_targets <- c("paleo_raw", "paleo_with_ids", "paleo_merged_counts")
  
  tar_visnetwork(
    targets_only = TRUE, 
    names = paleo_targets,
    label = c("time", "size")
  ) %>%
    visLegend(zoom = FALSE)
}, error = function(e) {
  cat("Paleo pipeline visualization not available.\n")
})
```

## ID Extraction and Sample Classification

```{r paleo-id-extraction}
tryCatch({
  cat("Paleo ID Extraction Results:\n")
  cat("Raw microscopy lines:", nrow(paleo_raw), "\n")
  cat("Unique samples after ID extraction:", length(unique(paleo_with_ids$Sample_ID)), "\n")
  cat("Unique V_numbers:", length(unique(paleo_with_ids$V_number[!is.na(paleo_with_ids$V_number)])), "\n")
  cat("Unique LSPECs:", length(unique(paleo_with_ids$LSPEC[!is.na(paleo_with_ids$LSPEC)])), "\n\n")
  
  # Sample type distribution
  if ("sample_type" %in% names(paleo_with_ids)) {
    sample_types <- table(paleo_with_ids$sample_type, useNA = "ifany")
    cat("Sample Type Classification:\n")
    print(sample_types)
    cat("\n")
  }
  
  # Show example ID parsing
  sample_paleo <- head(paleo_with_ids[!is.na(paleo_with_ids$V_number), 
                                     c("Sample_ID", "V_number", "LSPEC", "sample_type")], 5)
  kable(sample_paleo, caption = "Example Paleo ID Parsing")
}, error = function(e) {
  cat("Paleo ID extraction data not available.\n")
})
```

## Microscopy Count Aggregation

```{r count-merging}
tryCatch({
  cat("Microscopy Count Merging Results:\n")
  cat("Input microscopy lines:", nrow(paleo_raw), "\n")
  cat("Output merged samples:", nrow(paleo_merged_counts), "\n")
  cat("Data reduction:", round((1 - nrow(paleo_merged_counts)/nrow(paleo_raw)) * 100, 1), "%\n\n")
  
  # Count distribution by microfossil type
  if ("Microfossil_Type" %in% names(paleo_merged_counts)) {
    type_dist <- paleo_merged_counts %>%
      count(Microfossil_Type, name = "Samples") %>%
      arrange(desc(Samples))
    
    cat("Microfossil Type Distribution:\n")
    kable(type_dist, caption = "Sample Counts by Microfossil Type")
  }
}, error = function(e) {
  cat("Count merging data not available.\n")
})
```

## Microfossil Distribution

```{r microfossil-plot, fig.height=6}
tryCatch({
  print(plot_microfossil_types)
}, error = function(e) {
  cat("Microfossil type plot not available.\n")
})
```

---

# Field Order Processing Pipeline

The field order pipeline processes stratigraphic and age data that links fish specimens to their geological context.

## Field Order Pipeline Flow

```{r fieldorder-pipeline-viz, fig.height=4}
tryCatch({
  fieldorder_targets <- c(
    "order_raw", "order_with_ids", "fieldorder_formatted",
    "fieldorder_duplicates_clean", "fieldorder_updated", "fieldorder_final"
  )
  
  tar_visnetwork(
    targets_only = TRUE, 
    names = fieldorder_targets,
    label = c("time", "size")
  ) %>%
    visLegend(zoom = FALSE)
}, error = function(e) {
  cat("Field order pipeline visualization not available.\n")
})
```

## Field Order Processing Summary

```{r fieldorder-summary}
tryCatch({
  summary_data <- fieldorder_processing_summary
  
  # Create summary table
  processing_table <- data.frame(
    Metric = c(
      "Input Records", "Final Complete Records", "Data Retention Rate",
      "Duplicate Conflicts", "Missing Data Records"
    ),
    Value = c(
      summary_data$record_counts$input_records,
      summary_data$record_counts$final_complete_records,
      paste0(summary_data$record_counts$data_retention_rate, "%"),
      summary_data$data_quality$duplicate_conflicts,
      summary_data$data_quality$missing_data_records
    )
  )
  
  kable(processing_table, caption = "Field Order Processing Summary")
  
  cat("\nProcessing Steps Completed:\n")
  cat("1.", summary_data$processing_steps$step1_formatting, "\n")
  cat("2.", summary_data$processing_steps$step2_duplicates, "\n") 
  cat("3.", summary_data$processing_steps$step3_updates, "\n")
  cat("4.", summary_data$processing_steps$step4_missing_data, "\n")
}, error = function(e) {
  cat("Field order processing summary not available.\n")
})
```

## Stratigraphic and Age Distributions

```{r fieldorder-distributions, fig.height=8}
tryCatch({
  # Create distribution plots
  p1 <- stratigraphic_distribution_plot(fieldorder_final)
  p2 <- age_distribution_plot(fieldorder_final)
  
  # Print plots
  print(p1)
  print(p2)
}, error = function(e) {
  cat("Field order distribution plots not available.\n")
})
```

---

# Stratigraphic Integration

The rioja integration pipeline combines stickleback paleo-ecological data with stratigraphic information to create summary tables and stratigraphic plots for evolutionary analysis.

## LSPEC Integration Results

```{r lspec-integration-summary}
tryCatch({
  integration_summary <- lspec_integration_check$summary
  
  cat("LSPEC Integration Summary:\n")
  cat("Paleo stickleback LSPECs:", integration_summary$paleo_lspecs_total, "\n")
  cat("Field order LSPECs available:", integration_summary$fieldorder_lspecs_total, "\n")
  cat("Successfully linked LSPECs:", integration_summary$linkable_lspecs, "\n")
  cat("Missing from field order:", integration_summary$missing_from_fieldorder, "\n")
  cat("Integration success rate:", integration_summary$linkage_rate, "%\n\n")
  
  if (integration_summary$missing_from_fieldorder > 0) {
    cat("Missing LSPECs flagged and saved to 'data/flagged/paleo_lspecs_missing_fieldorder.csv'\n")
  }
}, error = function(e) {
  cat("LSPEC integration data not available.\n")
})
```

## Summary Table

```{r paleo-summary-table}
tryCatch({
  cat("Summary Table Dimensions:\n")
  cat("LSPECs (rows):", nrow(paleo_summary_table), "\n")
  cat("Total columns:", ncol(paleo_summary_table), "\n")
  
  # Count metadata vs taxa columns
  metadata_cols <- c("LSPEC", "YEAR", "CSTRAT", "ISTRAT", "INT", "Sample_ID", "V_number", "total_count", "n_taxa")
  available_metadata <- intersect(metadata_cols, names(paleo_summary_table))
  taxa_cols <- setdiff(names(paleo_summary_table), available_metadata)
  
  cat("Metadata columns:", length(available_metadata), "\n")
  cat("Taxa columns:", length(taxa_cols), "\n\n")
  
  # Show age range
  if ("YEAR" %in% names(paleo_summary_table)) {
    age_range <- range(paleo_summary_table$YEAR, na.rm = TRUE)
    cat("Age range:", age_range[1], "-", age_range[2], "years\n")
  }
  
  # Show preview
  preview_data <- paleo_summary_table[1:min(6, nrow(paleo_summary_table)), 
                                     1:min(10, ncol(paleo_summary_table))]
  kable(preview_data, caption = "Paleo Summary Table Preview (first 6 rows, 10 columns)")
}, error = function(e) {
  cat("Paleo summary table not available.\n")
})
```

## Rioja Stratigraphic Plot

```{r rioja-plot, fig.height=10, fig.width=12}
tryCatch({
  # Display plot summary
  if (!is.null(rioja_strat_plots$summary)) {
    summary <- rioja_strat_plots$summary
    cat("Rioja Plot Summary:\n")
    cat("Samples plotted:", summary$n_samples, "\n")
    cat("Taxa plotted:", summary$n_taxa, "\n")
    cat("Taxa removed (zero counts):", summary$taxa_removed, "\n")
    cat("Total specimens displayed:", summary$total_abundance, "\n\n")
  }
  
  # Recreate the plot for display
  if (!is.null(rioja_strat_plots$species_data) && !is.null(rioja_strat_plots$depth_data)) {
    rioja::strat.plot(
      d = rioja_strat_plots$species_data,
      yvar = rioja_strat_plots$depth_data,
      title = "Microfossil Stratigraphic Distribution",
      ylabel = "Age (years)",
      srt.xlabel = 45,
      mgp = c(3, 1, 0)
    )
  }
}, error = function(e) {
  cat("Rioja stratigraphic plot not available. Run the rioja pipeline targets to generate plots.\n")
})
```

---

# Community Ecology Analysis

The community ecology pipeline applies modern community ecology metrics to the integrated paleo-ecological data, providing quantitative insights into how microfossil communities changed through geological time.

## Community Metrics Summary

```{r community-metrics-summary}
tryCatch({
  cat("Community Metrics Summary:\n")
  cat("Time bins analyzed:", nrow(community_metrics), "\n")
  
  if ("YEAR" %in% names(community_metrics)) {
    cat("Temporal range:", min(community_metrics$YEAR), "-", max(community_metrics$YEAR), "years\n")
  }
  if ("CSTRAT" %in% names(community_metrics)) {
    cat("Stratigraphic range:", min(community_metrics$CSTRAT), "-", max(community_metrics$CSTRAT), "cm\n")
  }
  
  # Create summary table
  metrics_summary <- community_metrics %>%
    select(where(is.numeric), -any_of(c("CSTRAT", "YEAR"))) %>%
    summarise_all(list(
      Min = ~min(., na.rm = TRUE),
      Max = ~max(., na.rm = TRUE), 
      Mean = ~mean(., na.rm = TRUE),
      Median = ~median(., na.rm = TRUE)
    )) %>%
    tidyr::pivot_longer(everything(), names_to = "Metric_Stat", values_to = "Value") %>%
    tidyr::separate(Metric_Stat, into = c("Metric", "Statistic"), sep = "_(?=[^_]*$)") %>%
    tidyr::pivot_wider(names_from = Statistic, values_from = Value) %>%
    mutate(across(where(is.numeric), ~round(., 3)))
  
  kable(metrics_summary, caption = "Community Metrics Summary Statistics")
}, error = function(e) {
  cat("Community metrics data not available.\n")
})
```

## Community Trends Through Time

```{r community-trends-plot, fig.height=12, fig.width=10}
tryCatch({
  print(trends_plot)
}, error = function(e) {
  cat("Community trends plot not available.\n")
})
```

## Pairwise Community Turnover

```{r turnover-analysis}
tryCatch({
  cat("Pairwise Turnover Analysis:\n")
  cat("Method:", turnover_matrix$method, "\n")
  cat("Time bins:", turnover_matrix$summary$n_time_bins, "\n")
  cat("Pairwise comparisons:", turnover_matrix$summary$n_comparisons, "\n")
  cat("Turnover range:", round(turnover_matrix$summary$min_turnover, 3), "-", 
      round(turnover_matrix$summary$max_turnover, 3), "\n")
  cat("Mean turnover:", round(turnover_matrix$summary$mean_turnover, 3), "\n\n")
}, error = function(e) {
  cat("Turnover analysis data not available.\n")
})
```

```{r turnover-heatmap-plot, fig.height=8, fig.width=10}
tryCatch({
  print(turnover_heatmap)
}, error = function(e) {
  cat("Turnover heatmap not available.\n")
})
```

---

# Pipeline Summary

## Data Transformation Overview

```{r transformation-summary}
tryCatch({
  summary_data <- data.frame(
    Dataset = c("Morphology", "Paleo-Ecology"),
    Input_Records = c(nrow(morph_raw), nrow(paleo_raw)),
    Output_Records = c(nrow(morph_final), nrow(paleo_merged_counts)),
    Input_Units = c(length(unique(morph_raw$ID)), length(unique(paleo_raw$Sample_ID))),
    Output_Units = c(length(unique(morph_final$fish_id)), length(unique(paleo_merged_counts$Sample_ID))),
    Data_Reduction = c(
      paste0(round((1 - nrow(morph_final)/nrow(morph_raw)) * 100, 1), "%"),
      paste0(round((1 - nrow(paleo_merged_counts)/nrow(paleo_raw)) * 100, 1), "%")
    )
  )
  
  kable(summary_data, caption = "Pipeline Transformation Summary")
}, error = function(e) {
  cat("Summary data not available.\n")
})
```

## Key Quality Control Actions

**Morphology Pipeline**:
- Standardized specimen IDs and extracted fish/geological identifiers
- Resolved missing scale bar issues through colleague collaboration
- Applied biological logic to distinguish evolutionary loss from preservation artifacts  
- Flagged and removed fish with conflicting part/counterpart measurements
- Merged clean part/counterpart data prioritizing higher-quality measurements

**Paleo-Ecology Pipeline**:
- Extracted and standardized sample identifiers and classifications
- Aggregated individual microscopy line counts into geological sample totals
- Preserved complete taxonomic resolution while reducing data volume

**Field Order Pipeline**:
- Standardized LSPEC identifiers for cross-dataset linking
- Resolved duplicate records through conflict detection and clean merging
- Incorporated additional stratigraphic data where available
- Flagged incomplete records for manual review

Both datasets are now ready for integrated evolutionary analysis using LSPEC identifiers to link fish morphology with their paleo-environmental context.

</div>

---

*Report generated on `r Sys.Date()` using targets pipeline*